Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-27 Manuscript/manuscript.Rmd
methods edits

Architectural proteins tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To measure the pairwise correlation of each cell line specific annotation, we used the Jaccard Index. The Jaccard Index measures the ratio of the number of intersecting base pairs between two sets to the number of base pairs in the union, subtracted by the intersection [@favorov2012exploring]. The quantity ranges from 0.0 to 1.0, where 0.0 represents no overlap (no correlation) and 1.0 represent complete overlap (complete correlation).

In order to suitably reduce the feature space while also maintaining the inclusion of possibly colocalized annotations, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. Given that our approach is chromosome specific, we aggregated the predictive importance of the union of the optimal set of features across chromosome using the mean decrease in node impurity. We then performed hierarchical clustering to obtained the most common and top ranked annotations responsible for TAD boundary prediction.

@@

Random forest classification models were built to compare model performances between combinations of predictor types and re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, n=500. All random forest algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. Once the model was implemented, it was validated on the test set. The performance of each model was compared using the composite metric, Matthews Correlation Coefficient (MCC). The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction. The MCC is defined as

@@

## USING OPTIMALLY BUILT CLASSIFIER TO PREDICT BOUNDARIES AT FINER RESOLUTION

[needs editting]

>>>>

@@

Random forest classification models were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. To reduce bias due to random dataset generation, 10-fold cross validation was used to tune the number of annotations, *p*, to consider at each node. The number of trees that were aggregated for each random forest model was set to the default, n=500. All random forest algorithms were implemented in R using the *caret* package version 6.0 [@kuhn2012caret]. The performance of each model was compared using Matthews Correlation Coefficient (MCC), defined as <!--The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of 1 represents a perfect prediction, while a value of -1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction. The MCC is defined as-->

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. This can contribute to high collinearity which can result in overfitting a predictive model. To measure the pairwise correlation of each cell line specific annotation, we used the Jaccard Index. The Jaccard Index measures the ratio of the number of intersecting base pairs between two sets to the number of base pairs in the union, subtracted by the intersection [@favorov2012exploring]. The quantity ranges from 0.0 to 1.0, where 0.0 represents no overlap (no correlation) and 1.0 represent complete overlap (complete correlation).

In order to suitably reduce the feature space and prevent overfitting, we implemented recursive feature elimination to estimate the optimal number of necessary features to include in the model. The best performing subset was identified, ranging from 2 to the maximum number of features by powers of 2. Given that our approach is chromosome specific, we aggregated the predictive importance of the union of the optimal set of features across chromosome using the mean decrease in node impurity. We then performed hierarchical clustering to obtained the most common and top ranked annotations responsible for TAD boundary prediction.

## USING OPTIMALLY BUILT RANDOM FOREST CLASSIFIERS FOR MORE PRECISE TAD BOUNDARY PREDICTION

Once we established our optimally built and feature reduced random forest classifier, we aimed to leverage the enrichment of known annotations to predict TAD boundaries at base pair resolution. We established a *predicted TAD boundary region* as a collection of adjacent base pairs that were given a probability of 1 by our model in being a predicted TAD boundary. To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary region*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. Lastly, the midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a predicted TAD boundary.
